Verified Nanopasses for Compiling Conditionals

### 2025-08-12 paper.tex
progress

The third language replaces the \lstinline{Let} bindings with
assignment expressions but retains the structured control-flow and
expression-oriented nature of the language. This language brings us
closer to x86 because variables are represented by registers and stack
locations, all of which are mutable. The abstract syntax for
assignment is \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the
right-hand side and $e_2$ is an expression to be executed after the
assignment (reminiscent of the body of a \lstinline{Let}).  The
\lstinline{Program} AST node consists of the number of variables and
an expression.

@@

The fourth language is similar to the C programming language in that
it supports \texttt{goto} and separates expressions from statements.
However, to get us closer to x86, the condition of an \lstinline{if}
statement is restricted to an equality test. This form of
\lstinline{if} can be easily translated into a sequence of three x86
instructions: \lstinline{cmpq}, \lstinline{je}, and \lstinline{jmp}.
However, compiling from the previous IR to this one is non-trivial
because the previous IR allowed arbitrarily nested expressions to
appear in the condition of an \lstinline{if}, including other
\lstinline{if}'s!

@@

We choose a big-step semantics for C-like language because the
semantics is no longer structurally recursive and furthermore,
programs are no longer guaranteed to terminate.\footnote{The programs
in the image of Explicate Control are guaranteed to terminate, but
Agda doesn't know that.}  So encoding the semantics as a function in
Agda would come with complications (such as using a gas parameter)
that we side-step by using a big-step semantics. Compared to the
semantics for the imperative IR, this semantics includes a parameter
for the control flow graph.

@@

The final intermediate language is a subset of x86 assembly language,
except for the addition of variables. A complete compiler would
include a few more passes to perform register allocation and deal with
idiosyncrasies of x86, such as only allowing at most one instruction
argument to access memory.

@@

The meaning of each instruction is given by the following relation.
The Boolean in this relation specifies whether the instruction
falls-through to the next instruction or not (because it jumped).

@@

\noindent The meaning of a sequence of instructions is given the
following big-step relation.

\begin{lstlisting}
data _,_$⊩$_$⇓$_,_ : StateX86 $→$ List Block $→$ List Inst
                     $→$ StateX86 $→$ $𝔹$ $→$ Set 
\end{lstlisting}

This big-step treatment of the semantics of instruction sequences is
unusual, so here are the three rules.

\begin{gather*}
\inferrule{}{s , B ⊩ [] ⇓ s , \mathsf{true}} \\[1ex]
\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{true} \\
           s′ , B ⊩ is ⇓ s″ , b}
          {s , B ⊩ i ∷ is ⇓ s′ , b}\\[1ex]
\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{false}}
          {s , B ⊩ i ∷ is ⇓ s′ , b}
\end{gather*}

The meaning of a whole program is given by the following relation.
\begin{lstlisting}
interp-x86-var : X86Var $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

We now proceed to discuss the compiler and its proof of correctness.

@@

The compiler is simply the composition of the four passes,
which we detail in the next four subsections.

@@

The Remove Complex Operands pass is responsible for making sure that
the arguments of primitive operations, such as \lstinline{Sub} and
\lstinline{Eq}, are atomic expression (variables or literals). It does
so by insertion \lstinline{let} bindings for new temporary variables.
In the running example (figure~\ref{fig:running-example-monadic}), the
subexpressions \lstinline{read} and \lstinline{y - 2} are both
arguments of primitive operations, so the Remove Complex Operands pass
binds them to the temporary variables \lstinline{t1} and
\lstinline{t2}, respectively.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else (let t1 = read in t1 == 1))
    then (y - 1)
    else (let t2 = y - 2 in t2 - x)
\end{lstlisting}
  \hrule
\caption{Example after Remove Complex Operands, in
  monadic normal form.}
\label{fig:running-example-monadic}
\end{figure}

@@

\begin{lstlisting}
rco (Sub e$₁$ e$₂$)
    with rco e$₁$ | rco e$₂$
... | m$₁$ | m$₂$
    with atomic? m$₁$ | atomic? m$₂$
... | yes (atomic a$₁$) | yes (atomic a$₂$) =
      Sub a$₁$ a$₂$
... | no cmplx$₁$ | yes (atomic a$₂$) =
      Let (rco e$₁$) (Sub (Var zero) (shift-atm a$₂$ 0))
... | yes (atomic a$₁$) | no cmplx$₂$ =
      Let (rco e$₂$) (Sub (shift-atm a$₁$ 0) (Var 0))
... | no cmplx$₁$ | no cmplx$₂$ = 
      Let m$₁$
        (Let (shift-mon m$₂$ 0)
          (Sub (Var 1) (Var 0)))
\end{lstlisting}

@@

The Lift Locals pass is responsible for replacing \lstinline{let}
expressions with assignment statements. In the output language,
variables are declared at the top of the program.  Here is the result
of Lift Locals on the running example. The output of Lift Locals on
the running example is shown in
figure~\ref{fig:running-example-structured-imperative}.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    x := read;
    y := read;
    if (if x == 0
        then true
        else (t1 := read; t1 == 1)
    then (y - 1)
    else (t2 := y - 2; t2 - x)
\end{lstlisting}
  \hrule
  \caption{Example after Lift Locals, in an imperative IR.}
  \label{fig:running-example-structured-imperative}
\end{figure}
The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns a number $n$ that counts the number of
\lstinline{let} bindings in the expression, and the translated
expression.

\begin{lstlisting}
lift-locals : Mon $→$ Imp-Prog
lift-locals m
    with lift-locals-mon m
... | n , e = Program n e    
\end{lstlisting}

\noindent The case of \lstinline{lift-locals-mon} for \lstinline{Let}
is the most interesting. One can imagine that the result of
\lstinline{lift-locals-mon m$₁$} is a sequence of $i$ imaginary
\lstinline{let}'s surrounding \lstinline{e$₁$}. Similarly for 
\lstinline{lift-locals-mon m$₁$}.
The translation result for this \lstinline{Let} should be
a sequence of $i + j + 1$ imaginary
\lstinline{let}'s, followed by an assignment
of \lstinline{e$₁$} to the appropriate variable, and then the
code for \lstinline{e$₂$}. However, we need to shift
the variables in \lstinline{e$₁$} up by $j + 1$ because
we are moving all of the $j$ \lstinline{let}'s for
\lstinline{e$₂$} above \lstinline{e$₁$}, as well as the
current \lstinline{let}. Thus, we need another function
for shifting de Bruijn indices by an arbitrary amount.

@@

The variables in \lstinline{e$₂$} also need to be shifted.  They are
shifted up by $i$ because we are moving all of the $i$
\lstinline{let}'s for \lstinline{e$₁$} above
\lstinline{e$₂$}. Regarding the variable that serves as the left-hand
side of the assignment, its de Bruijn index is $i + j$ because the
original index for this \lstinline{Let} was $0$ but we have moved $i +
j$ other \lstinline{let}s above this variable occurrence.

\begin{lstlisting}
lift-locals-mon (Let m$₁$ m$₂$)
    with lift-locals-mon m$₁$
... | i , e$₁$
    with lift-locals-mon m$₂$
... | j , e$₂$ =
    let e$′₁$ = shifts-imp-exp e$₁$ 0 (suc j) in
    let e$′₂$ = shifts-imp-exp e$₂$ j i in
    suc (i + j) , Assign (i + j) e$′₁$ e$′₂$
\end{lstlisting}

@@

This theorem is a corollary of the following main lemma, which is
proved by structural induction on the expression \lstinline{m}.
Considerable care is needed in the statement of this lemma so that the
induction hypotheses can be used. Recall that when interpreting an
expression, the environment $ρ$ provides the meaning for the
\emph{free variables} in the expression.  As the interpreter descends
under \lstinline{let} bindings, it pushes values on the front of the
environment. However, after Lift Locals, all of the variables are
bound at the top, so the initial environment won't be empty, but
instead will have entries for all of the variables (we arbitrarily
initialize them to $0$). So in general, we can separate the
environment into two halves, $ρ₁$ and $ρ₂$, where $ρ₂$ is for the free
variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
free variables in the result expression \lstinline{e} of
\lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
number returned by \lstinline{lift-locals-mon m}. Also, note that
because \lstinline{Mon} is a pure language, \lstinline{e} contains a
single assignment to each variable, so $ρ₂$ is unchanged by the
execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
are assigned to for the first time.)

@@

The main lemma relies on more lemmas about shifting de Bruijn indices
that generalize our previous ones to arbitrary amounts of shifting.

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    block_0:
       x := read;
       y := read;
       if (x == 0) goto block_1; else goto block_3;
    block_1:
        goto block_2;
    block_2:
        return y - 1;
    block_3:
       t1 := read;
       if (t1 == 1) goto block_4; else goto block_5;
    block_4:
        goto block_2;
    block_5:
        goto block_6;
    block_6:
        t2 := y - 2;
        return t2 - x;
\end{lstlisting}
  \hrule
  \caption{Example after Explicate Control, with control flow
    made explicit with goto.}
\label{fig:running-example-goto}
\end{figure}

@@

All three functions will be building a control flow graph, which needs
to be threaded into and out of these functions, so they use the
following \lstinline{Build} monad:

@@

\noindent The following four functions make up the Explicate Control
pass. Next we discuss select cases of these functions.

@@

The case for conditional expressions is interesting: we do not
directly generate an assignment statement. The two branches,
\lstinline{e$₂$} and \lstinline{e$₃$}, are in assignment position,
so we recursively compile them with \lstinline{explicate-assign}.
They both have the same continuation \lstinline{c} and the first rule
of compilation is that you cannot duplicate code.\footnote{GHC
violated this rule a decade ago:
\url{https://stackoverflow.com/questions/35815503/what-ghc-optimization-is-responsible-for-duplicating-case-expressions}.}
Thus, we add \lstinline{c} to the control-flow graph and
instead use a \lstinline{Goto} to its label as the continuation
for \lstinline{e$₂$} and \lstinline{e$₃$}.

@@

For the finale of the Explicate Control pass, we discuss the
translation of conditionals in \lstinline{explicate-pred}.  The two
branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
position, so we compile them with recursive calls to
\lstinline{explicate-pred} to obtain \lstinline{c$₂$} and
\lstinline{c$₃$}. However, they share the same two continuations,
\lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
to avoid code duplication. Finally, we compile \lstinline{e$₁$} with
\lstinline{explicate-pred}, passing \lstinline{c$₂$} and
\lstinline{c$₃$} as the continuations.

>>>>

The third language replaces the \lstinline{Let} bindings with
assignment expressions but retains the structured control-flow and
expression-oriented nature of the language. This language brings us
closer to x86 where variables are replaced by registers and stack
locations, all of which are mutable. The abstract syntax for
assignment is \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the
right-hand side and $e_2$ is an expression to be executed after the
assignment (reminiscent of the body of a \lstinline{Let}).  The
\lstinline{Program} AST node consists of the number of variables and
an expression.

@@

The fourth language is similar to the C programming language in that
it supports \texttt{goto} and separates expressions from statements.
However, to get us closer to x86, the condition of an \lstinline{if}
statement is restricted to an equality test. This form of
\lstinline{if} can be easily translated into a sequence of three x86
instructions: \lstinline{cmpq}, \lstinline{je}, and \lstinline{jmp}.
However, compiling from the previous IR to this one is non-trivial
because the previous IR allowed arbitrarily nested expressions to
appear in the condition of an \lstinline{if}, including other
\lstinline{if}s!

@@

We choose a big-step semantics for the C-like language because the
semantics is no longer structurally recursive and furthermore,
programs are no longer guaranteed to terminate.\footnote{The programs
in the image of Explicate Control are guaranteed to terminate, but
Agda doesn't know that.}  So encoding the semantics as a function in
Agda would come with complications (such as using a gas parameter)
that we side-step by using a big-step semantics. Compared to the
semantics for the imperative IR, this semantics includes a parameter
for the control flow graph.

@@

The final intermediate language is a subset of x86 assembly language,
except for the addition of variables. The compiler in \emph{Essentials
of Compilation} includes a few more passes to perform register
allocation and deal with the idiosyncrasies of x86, such as only
allowing at most one instruction argument to access memory.

@@

The meaning of each instruction is given by the following relation.
The Boolean in this relation specifies whether the instruction
falls through to the next instruction or not (because it jumped).

@@

The meaning of a sequence of instructions is given the
following big-step relation.

\begin{lstlisting}
data _,_$⊩$_$⇓$_,_ : StateX86 $→$ List Block $→$ List Inst
                     $→$ StateX86 $→$ $𝔹$ $→$ Set 
\end{lstlisting}
This big-step treatment of instruction sequences is unusual, so here
are the three rules.
\begin{gather*}
\inferrule{}{s , B ⊩ [] ⇓ s , \mathsf{true}} \\[1ex]
\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{true} \\
           s′ , B ⊩ is ⇓ s″ , b}
          {s , B ⊩ i ∷ is ⇓ s′ , b}\\[1ex]
\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{false}}
          {s , B ⊩ i ∷ is ⇓ s′ , \mathsf{false}}
\end{gather*}
The meaning of a program is given by the following relation.
\begin{lstlisting}
interp-x86-var : X86Var $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

Next we discuss the compiler and its proof of correctness.

@@

The compiler is the composition of the four passes, which we detail in
the next four subsections.

@@

The Remove Complex Operands pass is responsible for making sure that
the arguments of primitive operations, such as \lstinline{Sub} and
\lstinline{Eq}, are atomic expression (variables or literals). It does
so by inserting \lstinline{let} bindings for new temporary variables.
In the running example (figure~\ref{fig:running-example-monadic}), the
subexpressions \lstinline{read} and \lstinline{(y - 2)} are both
arguments of primitive operations, so the Remove Complex Operands pass
binds them to the temporary variables \lstinline{t1} and
\lstinline{t2}, respectively.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    let x = read in
    let y = read in
    if (if x == 0
        then true
        else (let t1 = read in t1 == 1))
    then (y - 1)
    else (let t2 = y - 2 in t2 - x)
\end{lstlisting}
  \hrule
\caption{\small Result of Remove Complex Operands on the example.}
\label{fig:running-example-monadic}
\end{figure}

@@

\begin{lstlisting}
rco (Sub e$₁$ e$₂$)
    with rco e$₁$ $|$ rco e$₂$
... $|$ m$₁$ $|$ m$₂$
    with atomic? m$₁$ $|$ atomic? m$₂$
... $|$ yes (atomic a$₁$) $|$ yes (atomic a$₂$) =
      Sub a$₁$ a$₂$
... $|$ no cmplx$₁$ $|$ yes (atomic a$₂$) =
      Let (rco e$₁$) (Sub (Var zero) (shift-atm a$₂$ 0))
... $|$ yes (atomic a$₁$) $|$ no cmplx$₂$ =
      Let (rco e$₂$) (Sub (shift-atm a$₁$ 0) (Var 0))
... $|$ no cmplx$₁$ $|$ no cmplx$₂$ = 
      Let m$₁$
        (Let (shift-mon m$₂$ 0)
          (Sub (Var 1) (Var 0)))
\end{lstlisting}

@@

The Lift Locals pass is responsible for replacing \lstinline{let}
expressions with assignment statements. In the output language,
variables are declared at the top of the program.  The result of Lift
Locals on the running example is shown in
figure~\ref{fig:running-example-structured-imperative}.

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    x := read;
    y := read;
    if (if x == 0
        then true
        else (t1 := read; t1 == 1)
    then (y - 1)
    else (t2 := y - 2; t2 - x)
\end{lstlisting}
  \hrule
  \caption{Result of Lift Locals on the example.}
  \label{fig:running-example-structured-imperative}
\end{figure}
The \lstinline{lift-locals} function delegates to the recursive
function \lstinline{lift-locals-mon}, where all the work is done.
That function returns (1) a number $n$ that counts the number of
\lstinline{let} bindings that were in the expression and (2) the
translation of the expression.

\begin{lstlisting}
lift-locals : Mon $→$ Imp-Prog
lift-locals m
    with lift-locals-mon m
... $|$ n , e = Program n e    
\end{lstlisting}

\noindent The case of \lstinline{lift-locals-mon} for \lstinline{(Let m$_1$ m$_2$)}
is the most interesting. Think of the result of
\lstinline{lift-locals-mon m$₁$} as a sequence of $i$ imaginary
\lstinline{let}'s surrounding \lstinline{e$₁$}. Similarly for
\lstinline{lift-locals-mon m$₁$}.  The translation result for
\lstinline{(Let m$_1$ m$_2$)} should be a sequence of $i + j + 1$ imaginary
\lstinline{let} bindings followed by an assignment of \lstinline{e$₁$} to
the appropriate variable, and then the code for
\lstinline{e$₂$}. However, we need to shift the variables in
\lstinline{e$₁$} up by $j + 1$ because we are moving all of the $j$
\lstinline{let} bindings for \lstinline{e$₂$} above \lstinline{e$₁$}, as well
as the current \lstinline{let}. Thus, we need another function for
shifting de Bruijn indices by an arbitrary amount.

@@

The variables in \lstinline{e$₂$} also need to be shifted.  They are
shifted up by $i$ because we are moving all of the $i$
\lstinline{let} bindings for \lstinline{e$₁$} above
\lstinline{e$₂$}. Regarding the variable that serves as the left-hand
side of the assignment, its de Bruijn index is $i + j$ because the
original index for this \lstinline{Let} was $0$ but we have moved $i +
j$ other \lstinline{let}s above this variable occurrence.

\begin{lstlisting}
lift-locals-mon (Let m$₁$ m$₂$)
    with lift-locals-mon m$₁$
... $|$ i , e$₁$
    with lift-locals-mon m$₂$
... $|$ j , e$₂$ =
    let e$′₁$ = shifts-imp-exp e$₁$ 0 (suc j) in
    let e$′₂$ = shifts-imp-exp e$₂$ j i in
    suc (i + j) , Assign (i + j) e$′₁$ e$′₂$
\end{lstlisting}

@@

This theorem is a corollary of the following main lemma, which is
proved by structural induction on the expression \lstinline{m}.
Considerable care is needed in the statement of this lemma so that the
induction hypotheses can be used. Recall that when interpreting an
expression, the environment $ρ$ provides the meaning for the
\emph{free variables} in the expression.  As the interpreter descends
under \lstinline{let} bindings, it pushes values on the front of the
environment. However, after Lift Locals, all of the variables are
bound at the top, so the initial environment won't be empty, but
instead will have entries for all of the variables (we arbitrarily
initialize them to $0$). So in general, we can separate the
environment into two halves, $ρ₁$ and $ρ₂$, where $ρ₂$ is for the free
variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
free variables in the result expression \lstinline{e} of
\lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
number returned by \lstinline{lift-locals-mon m}. Also, note that
because \lstinline{Mon} is a pure language, \lstinline{e} contains a
single assignment to each variable, so $ρ₂$ is unchanged by the
execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
are assigned for the first time.)

@@

The main lemma relies on another lemma about shifting de Bruijn
indices that generalizes our previous one to arbitrary amounts of
shifting.

@@

\begin{figure}[tp]
  \hrule
\begin{lstlisting}
    var x, y, t1, t2;
    block_0:
       x := read;
       y := read;
       if (x == 0) goto block_1; else goto block_3;
    block_1:
        goto block_2;
    block_2:
        return y - 1;
    block_3:
       t1 := read;
       if (t1 == 1) goto block_4; else goto block_5;
    block_4:
        goto block_2;
    block_5:
        goto block_6;
    block_6:
        t2 := y - 2;
        return t2 - x;
\end{lstlisting}
  \hrule
  \caption{Example after Explicate Control, where control flow
    is made explicit with goto.}
\label{fig:running-example-goto}
\end{figure}

@@

All three functions build a control flow graph, which needs to be
threaded into and out of these functions, so they use the following
\lstinline{Build} monad:

@@

\noindent The following four functions make up the Explicate Control
pass. Next we discuss selected cases of these functions.

@@

The case for conditional expressions is interesting: we do not
directly generate an assignment statement. The two branches,
\lstinline{e$₂$} and \lstinline{e$₃$}, are in assignment position,
so we recursively compile them with \lstinline{explicate-assign}.
They both have the same continuation \lstinline{c} and the first rule
of compilation is that you cannot duplicate code.\footnote{GHC
violated this rule a decade ago:
\url{https://stackoverflow.com/questions/35815503/what-ghc-optimization-is-responsible-for-duplicating-case-expressions}.}
Thus, we add \lstinline{c} to the control-flow graph and
instead use a \lstinline{Goto} to the label for \lstinline{c}
as the continuation for \lstinline{e$₂$} and \lstinline{e$₃$}.

@@

For the finale of the Explicate Control pass, we discuss the
translation of the conditional \lstinline{(If e$₁$ e$₂$ e$₃$)}
in \lstinline{explicate-pred}.  The two
branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
position, so we compile them with recursive calls to
\lstinline{explicate-pred} to obtain \lstinline{c$₂$} and
\lstinline{c$₃$}. However, they share the same two continuations,
\lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
to avoid code duplication. We compile \lstinline{e$₁$} with
\lstinline{explicate-pred}, passing \lstinline{c$₂$} and
\lstinline{c$₃$} as the continuations.
